Detection and replication of epistasis influencing transcription in humans

### 2013-12-10 docs/manuscript/methods.tex
still have a few words to cut

>>>>

\subsection{Limitations of this type of comparison}

Though it is useful to compare the relative variances of epistatic and additive effects, it must be stressed that our results here are approximations that are very limited by the study design. We estimate that additive effects explain approximately 10 times more variance than epistatic effects, but this could be an overestimate or an underestimate due to a number of different caveats. Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. Thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance. And forthly, the extent of winner's curse in estimation of effect sizes may differ between the two studies.

### 2013-12-10 docs/manuscript/shorter.tex
still have a few words to cut

% Epistasis is the phenomenon whereby a polymorphism's effect on a trait depends on other polymorphisms present in the genome. The extent to which epistasis influences complex traits \cite{Carlborg2004} and contributes to their variation \cite{Hill2008a, Crow2010} is a fundamental question in evolution and human genetics. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and examples have been reported in other species \cite{Carlborg2006}, few convincing examples with independent replication exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its low incidence in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but an alternative view is that it has previously been too technically challenging to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, many instances of epistasis are found between common single nucleotide polymorphisms (SNPs). In a cohort of 846 individuals with data on 7339 gene expression levels in peripheral blood, we found 501 significant pairwise epistatic interactions between common SNPs acting on the expression levels of 238 genes ($p < 2.91 \times 10^{-16}$). Replication of these interactions in two independent data sets \cite{Metspalu2004, Fehrmann2011} showed both concordance of direction of epistatic effects ($p = 5.56 \times 10^{-31}$) and enrichment of interaction $p$-values, with 30 being significant at a conservative threshold of $p < 0.05 / 501$. There was evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known physical chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Epistatic networks of three SNPs or more influence the expression levels of 129 genes, whereby one \emph{cis}-acting SNP is modulated by several \emph{trans}-acting SNPs. For example MBNL1 is influenced by an additive effect at rs13069559 which itself is masked by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first evidence for multiple instances of segregating common polymorphisms interacting to influence human traits.

@@

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. Thus, when combined with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants. Because our focus was to ascertain the extent to which instances of epistasis arises from natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes and like most complex diseases, these expression traits are typically heritable \cite{Schadt2003}. But unlike complex diseases, genetic associations with gene expression commonly have very large effect sizes that explain large proportions of the genetic variance\cite{Powell2013}, making them good candidates to search for epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions. First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood ($1.03 \times 10^{15}$ statistical tests, family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods). Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $0.05 \leq \alpha \leq 0.14$, Methods, Supplementary Figure \ref{fig:type1}). Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). We used strict quality control measures to avoid statistical associations being driven by technical artifacts (Methods). However it remains possible that unexplained technical artifacts may have led to the significant discovery interactions. Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/501$, Table \ref{tab:bonferroni}). To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A \times A$, $A \times D$, $D \times A$, $D \times D$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods). Sign concordance between the discovery and both replication datasets was observed in 22 out of the 30 significantly replicated interactions (expected value $= 7.5$ under the null hypothesis of no interactions, $p = 3.76 \times 10^{-8}$).

@@

% \subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for additive $\times$ additive genetic interactions \cite{Cordell2009}, but our analysis shows that this is unlikely to be the most effective strategy for its detection. The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \cite{Powell2013} (439 out of 501, Methods). Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects. Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst additive $\times$ additive, additive $\times$ dominance, dominance $\times$ additive and dominance $\times$ dominance at the discovery stage ($p = 0.22$ for departure from expectation). This is not surprising because the patterns of epistasis used for statistical decomposition (\emph{i.e.} $A \times A$, $A \times D$, $D \times A$, $D \times D$) are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.

@@

% \subsection{Contribution of epistatic variance}
Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question. In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance. Taking results from our previously published eQTL \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater. Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.

In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Taking all additive effects detected in Powell \emph{et al} (2012) that have additive variance explaining $2.1\%$ or greater of phenotypic variance, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$. By contrast, the estimated epistatic variance from the interacting SNPs detected in this study on average explain a total of $0.22\%$ of phenotypic variance, approximately ten times lower than the estimated additive variance. There are several caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different minimum variance thresholds, and our estimate is determined by the threshold used. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. Thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance. And forthly, the extent of winner's curse in estimation of effect sizes may differ between the the two studies.

>>>>

@@

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. Thus, with small genetic effect sizes, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes among common variants. Because our focus was to ascertain the extent to which instances of epistasis arises from natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes and like most complex diseases, these expression traits are typically heritable \cite{Schadt2003}. But unlike complex diseases, genetic associations with gene expression commonly have very large effect sizes that explain large proportions of the genetic variance\cite{Powell2013}, making them good candidates to search for epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we used a two stage approach to identify genetic interactions. First, we exhaustively test every pair of SNPs for pairwise effects against each of 7339 expression traits in peripheral blood ($1.03 \times 10^{15}$ statistical tests, family-wise error rate of 5\% corresponding to a significance threshold of $ p < 2.91 \times 10^{-16}$, Methods). Second, we filtered the SNP pairs from stage 1 on LD and genotype class counts, and tested the remaining pairwise effects for significant interaction terms and used a Bonferroni correction for multiple testing (estimated type 1 error rate $0.05 \leq \alpha \leq 0.14$, Methods, Supplementary Figure \ref{fig:type1}). Using this design we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). We used strict quality control measures to avoid statistical associations being driven by technical artifacts (Methods). However it remains possible that unexplained technical artifacts may have led to the significant discovery interactions. Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (5\% significance threshold $p < 0.05/501$, Table \ref{tab:bonferroni}). To quantify the similarity of GP maps between the independent datasets (Figure \ref{fig:gpmaps}) we decomposed the genetic effects of each of the SNP pairs into orthogonal additive, dominance and epistatic effects ($A1$, $A2$, $D1$, $D2$, $A1 \times A2$, $A1 \times D2$, $D1 \times A2$, $D1 \times D2$) and tested for concordance of the sign of the most signicant effect (Supplementary Table \ref{tab:sign}, Methods). Sign concordance between the discovery and both replication datasets was observed in 22 out of the 30 significantly replicated interactions (expected value $= 7.5$ under the null hypothesis of no interactions, $p = 3.76 \times 10^{-8}$).

@@

% \subsection{Patterns of epistasis}
Though seldom the focus of association studies, SNPs with known main effects are often tested for $A \times A$ genetic interactions \cite{Cordell2009}, but our analysis suggests this is unlikely to be the best strategy for its detection. The majority of our discovery interactions comprised of one SNP that was significantly associated with the gene expression level in the discovery dataset, and one SNP that had no previous association \cite{Powell2013} (439 out of 501, Methods). Only nine interactions were between SNPs that both had known main effects while 64 were between SNPs that had no known main effects. Additionally, we observed that the largest epistatic variance component for the 501 interactions was equally divided amongst $A \times A$, $A \times D$, $D \times A$ and $D \times D$ at the discovery stage ($p = 0.22$ for departure from expectation). This is not surprising because these patterns of epistasis used for statistical decomposition are simply convenient orthogonal parameterisations of a two locus model, and are not intended to model biological function \cite{Cockerham1954}.

@@

% \subsection{Contribution of epistatic variance}
Quantifying the importance of epistasis in complex traits in humans remains an open question. Here we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold, where the minimum estimated variance explained by the epistatic effects of any interaction was $2.1\%$ of phenotypic variance. Taking results from our previously published eQTL \cite{Powell2012} we calculated that 1848 of the 7339 gene expression levels analysed were influenced by additive effects where the estimated additive variance of a locus was $2.1\%$ or greater. Thus, we can infer that the number of instances of large additive effects is significantly greater than the number of instances of large epistatic effects.

In terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Taking all additive effects detected in Powell \emph{et al} (2012) that have additive variance explaining $2.1\%$ or greater of phenotypic variance, we calculated that the proportion of total phenotypic variance of all 7339 gene expression levels explained by additive effects alone was $2.16\%$. By contrast, the estimated epistatic variance from the interacting SNPs detected in this study on average explain a total of $0.22\%$ of phenotypic variance, approximately ten times lower than the estimated additive variance. There are several caveats to this comparison which we discuss in the Methods.
